Supporting Information:\\Black-swan events in animal populations

### 2016-06-24 text/supp.tex
More copyediting

\author{
Sean C. Anderson$^{1,2*}$ \and
Trevor A. Branch$^3$ \and
Andrew B. Cooper$^2$ \and
Nicholas K. Dulvy$^1$
}
\date{}

@@

\textsuperscript{2}School of Resource and Environmental Management, Simon
Fraser University, Burnaby, BC, V5A 1S6, Canada

\textsuperscript{3}School of Aquatic and Fishery Sciences, University of
Washington, Box 355020, Seattle, WA 98195, USA

@@

\section{Implementation of the skewed Student-t distribution in Stan}

The probability density for the skewed Student-t distribution
\citep{fernandez1998} equates to the following probability density as
implemented in \citet{king2012}:

\begin{equation}
  \mathrm{Skewed\,Student\mhyphen t}(x) =
\begin{dcases}
\frac{2}{(\gamma + 1/\gamma)}
         \mathrm{Student\mhyphen t}(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
       \frac{2}{(\gamma + 1/\gamma)}
         \mathrm{Student\mhyphen t}(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
\end{dcases}
\end{equation}

Here $\gamma$ represents the skewness parameter. The Student-t($\cdot$) portion
represents in order: the data point of interest, the heavy-tailedness
parameter, the location parameter, and the scale parameter as passed to the
standard Student-t probability density. This translates to the log probability
density:

\begin{equation}
\mathrm{Skewed\,Student\mhyphen t}(x) =
\begin{dcases}
  \log (2 \gamma) - \log(\gamma^2 + 1) +
  \log \mathrm{Student\mhyphen t}(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
  \log (2 \gamma) - \log(\gamma^2 + 1) +
  \log \mathrm{Student\mhyphen t}(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
\end{dcases}
\end{equation}

Since Stan only needs the probability density up to an additive
constant \citep{stan-manual2014}, we
can replace $\log(2 \gamma)$ in the above for $\log(\gamma)$. The final log
probability density for the skewed Student-t distribution implemented in
Stan\footnote{Credit to Stan developer Bob Carpenter:
\url{https://groups.google.com/d/msg/stan-users/jeZQnQRUsDs/BeqhicWm1GEJ}}
looks like the following:

@@

\textbf{Table~S1: Summary statistics for the filtered Global Population
Dynamics Database time series arranged by taxonomic class.} Columns are number
of populations, number of taxonomic orders, numbers of species, median time
series length, total number of interpolated time steps, total number of
substituted zeros, and total number of time steps.

@@

%\begin{helvetica}
\singlespacing
\begin{scriptsize}
\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{6.2cm}>{\RaggedRight}p{0.7cm}>{\RaggedRight}p{0.4cm}>{\RaggedRight}p{4.4cm}>{\RaggedRight}p{1.4cm}}
\caption{\textbf{Populations with a high probability of heavy-tailed dynamics in the base heavy-tailed Gompertz population dynamics model.} Shown are the log abundance time series, population descriptions, Global Population Dynamics Database Main IDs, citation for the data source or separate verification literature, a description of the cause of the black-swan events (if known), the probability of heavy tails as calculated by our model, and median estimate of $\nu$ from our model with 90\% quantile credible intervals indicated in parentheses. Red dots on the time series indicate downward black-swan events and blue values indicate upward black-swan events that have a $10^{-4}$ probability or less of occurring if the population dynamics were explained by a Gompertz model with normally distributed process noise with a standard deviation equal to the scale parameter in the fitted t distribution.}\\

@@

\caption{\textbf{Probability density of the Bayesian priors for the Gompertz models.} (\textbf{A}) per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$. (\textbf{B}) scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$ (\textbf{C}) Student-t distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$. (\textbf{D}) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$. Panel \textbf{C} also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$. The inset panel shows the same data but with a log-transformed x axis. Note that the base and weaker priors are relatively flat within the region of $\nu < 20$ with which we are concerned.}

@@

\caption{\textbf{Estimates of $\nu$ from alternative models plotted against the
    base Gompertz model estimates of $\nu$.} Alternative models are (\textbf{A})
    Ricker-logistic; (\textbf{B}) a Gompertz model with AR1 residual correlation;
    (\textbf{C}) modelling population growth rate with no density dependence; and
    (\textbf{D}) a Gompertz model with fixed magnitude of observation error. Shown
are medians of the posterior (dots) and 50\% credible intervals (segments). The
diagonal line indicates a one-to-one relationship. Different colours indicate
various taxonomic classes. The grey-shaded regions indicate regions of
disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics.
The Gompertz observation error model assumes a fixed standard deviation of
observation error of $0.2$ on a log scale. Models in \textbf{B} and \textbf{D}
that had not met convergence thresholds were run for four chains, 256,000 iterations (128,000 warm up), and retaining every 10th sample, after which the MCMC chains for one population in \textbf{B} and four populations in \textbf{D} had still not met convergence thresholds. These populations are excluded from their respective panels, but all populations with Pr($\nu < 10$) $> 0.5$ in the base model had converged and are shown.}

@@

\caption{\textbf{Estimates of $\nu$ from Gompertz models with alternative priors on $\nu$.} \textbf{A} shows a weaker prior on $\nu$, \textbf{B} shows a stronger prior on $\nu$, as illustrated in Fig.~1. Shown are medians of the posterior (dots) and 50\% credible intervals (segments). The diagonal line indicates a one-to-one relationship. Different colours indicate various taxonomic classes. The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics. In general, the estimates are nearly identical in cases where the data are informative about low values of $\nu$. When the data are less informative about low values of $\nu$, the prior can slightly pull the estimates of $\nu$ towards higher or lower values.}

@@

\caption{\textbf{Testing the ability to estimate $\nu$ (top panels) and the scale parameter of the process deviations (bottom panels) for a given number of samples (columns) drawn from a distribution with a given true $\nu$ value (rows).} The red lines indicate the true population value. When a small number of samples are drawn there may not be samples sufficiently far into the tails to recapture the true $\nu$ value; however, heavy tails are still distinguished from normal tails in most cases, even with only 25 or 50 samples.}

@@

\caption{\textbf{Simulation testing the Gompertz estimation model when the process deviation draws were chosen so that $\nu$ could be estimated close to the true value outside the full population model} (``effective $\nu$'' within a CV of 0.2 of specified $\nu$). Upper panels show the distribution of median $\widehat{\nu}$ across 20 simulation runs. Lower panels show the distribution of Pr($\nu < 10$) across 20 simulation runs. We ran the simulations across three population (``true'') $\nu$ values (3, 5, and $1\cdot 10^9$, i.e.\ approximately normal) and four scenarios: (1) 100 time steps and no observation error, (2) 50 time steps and no observation error, (3) 50 time steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4) 50 time steps with observation error in which the quantity of observation error was assumed known. Within each scenario the dots represent stochastic draws from the true population distributions combined with model fits. Underlayed boxplots show the median, interquartile range, and $1.5$ times the interquartile range.}

@@

\caption{\textbf{Potential covariates of heavy-tailed population dynamics (indicated by a high probability that $\nu < 10$).} Shown are (\textbf{A} to \textbf{C}) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (\textbf{D}) number of time steps, (\textbf{E}) body length, and (\textbf{F}) lifespan. For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-t process-noise distribution, $\lambda$ refers to the expected log abundance at the next time step at an abundance of one, $b$ refers to the density dependence parameter. Circles representing a few sharks, crustaceans, and gastropods are filled in white. Median and 90\% credible interval posterior predictions of a beta regression hierarchical model are shown in panels \textbf{A} and \textbf{D} where there was a high probability the slope coefficient was different from zero (Fig.~3B).}

@@

\caption{\textbf{Skewness and heavy-tailed parameter posteriors.} Top panel shows the shape of the skew-t distribution at different combinations of skewness ($\gamma$) and heavy-tailedness ($\nu$). Lower panel shows the posterior medians and 50\% credible intervals for the two parameters. The horizontal red line indicates the median ln(skewness) parameter across all populations.}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1.1\textwidth]{../analysis/heavy-skew-projections-ggplot}
\caption{\textbf{Projections of population abundance under skew-t (red) and
normally distributed process error models (blue) for populations with
Pr($\nu<10$) $>0.5$.} Numbers above panels indicate GPDD population IDs as
shown in table~S2. Horizontal axes indicates year before last
recorded abundance and a projection five years into the future. For the
projected region, solid lines represents the median projection and
dashed lines represent lower 99\% credible intervals. Vertical axes are
log-distributed. Assuming normal process error tends to underestimate risk of
low abundance if the process error is in fact heavy-tailed.}

>>>>

\author{
Sean C. Anderson$^{1*}$ \and
Trevor A. Branch$^2$ \and
Andrew B. Cooper$^3$ \and
Nicholas K. Dulvy$^1$
}
\date{}

@@

\textsuperscript{2}School of Aquatic and Fishery Sciences, University of
Washington, Box 355020, Seattle, WA 98195, USA

\textsuperscript{3}School of Resource and Environmental Management, Simon
Fraser University, Burnaby, BC, V5A 1S6, Canada

@@

\section{Implementation of the skewed Student-$t$ distribution in Stan}

The probability density for the skewed Student-$t$ distribution
\citep{fernandez1998} equates to the following probability density as
implemented in \citet{king2012}:

\begin{equation}
  \mathrm{Skewed\,Student}\mhyphen t(x) =
\begin{dcases}
\frac{2}{(\gamma + 1/\gamma)}
         \mathrm{Student}\mhyphen t(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
       \frac{2}{(\gamma + 1/\gamma)}
         \mathrm{Student}\mhyphen t(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
\end{dcases}
\end{equation}

Here $\gamma$ represents the skewness parameter. The Student-$t$($\cdot$) portion
represents in order: the data point of interest, the heavy-tailedness
parameter, the location parameter, and the scale parameter as passed to the
standard Student-$t$ probability density. This translates to the log probability
density:

\begin{equation}
\mathrm{Skewed\,Student}\mhyphen t(x) =
\begin{dcases}
  \log (2 \gamma) - \log(\gamma^2 + 1) +
  \log \mathrm{Student}\mhyphen t(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
  \log (2 \gamma) - \log(\gamma^2 + 1) +
  \log \mathrm{Student}\mhyphen t(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
\end{dcases}
\end{equation}

Since Stan only needs the probability density up to an additive
constant \citep{stan-manual2014}, we
can replace $\log(2 \gamma)$ in the above for $\log(\gamma)$. The final log
probability density for the skewed Student-$t$ distribution implemented in
Stan\footnote{Credit to Stan developer Bob Carpenter:
\url{https://groups.google.com/d/msg/stan-users/jeZQnQRUsDs/BeqhicWm1GEJ}}
looks like the following:

@@

Table~S1: Summary statistics for the filtered Global Population
Dynamics Database time series arranged by taxonomic class. Columns are number
of populations, number of taxonomic orders, numbers of species, median time
series length, total number of interpolated time steps, total number of
substituted zeros, and total number of time steps.

@@

%\begin{helvetica}
\singlespacing
\begin{scriptsize}
\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{6.2cm}>{\RaggedRight}p{0.7cm}>{\RaggedRight}p{0.4cm}>{\RaggedRight}p{4.4cm}>{\RaggedRight}p{1.4cm}}
\caption{Populations with a high probability of heavy-tailed dynamics in the base heavy-tailed Gompertz population dynamics model. Shown are the log abundance time series, population descriptions, Global Population Dynamics Database Main IDs, citation for the data source or separate verification literature, a description of the cause of the black-swan events (if known), the probability of heavy tails as calculated by our model, and median estimate of $\nu$ from our model with 90\% quantile credible intervals indicated in parentheses. Red dots on the time series indicate downward black-swan events and blue values indicate upward black-swan events that have a $10^{-4}$ probability or less of occurring if the population dynamics were explained by a Gompertz model with normally distributed process noise with a standard deviation equal to the scale parameter in the fitted $t$ distribution.}\\

@@

\caption{Probability density of the Bayesian priors for the Gompertz models. (\textbf{A}) per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$. (\textbf{B}) scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$ (\textbf{C}) Student-$t$ distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$. (\textbf{D}) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$. Panel \textbf{C} also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$. The inset panel shows the same data but with a log-transformed x axis. Note that the base and weaker priors are relatively flat within the region of $\nu < 20$ with which we are concerned.}

@@

\caption{Estimates of $\nu$ from alternative models plotted against the
    base Gompertz model estimates of $\nu$. Alternative models are (\textbf{A})
    Ricker-logistic; (\textbf{B}) a Gompertz model with AR1 residual correlation;
    (\textbf{C}) modelling population growth rate with no density dependence; and
    (\textbf{D}) a Gompertz model with fixed magnitude of observation error. Shown
are medians of the posterior (dots) and 50\% credible intervals (segments). The
diagonal line indicates a one-to-one relationship. Different colours indicate
various taxonomic classes. The grey-shaded regions indicate regions of
disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics.
The Gompertz observation error model assumes a fixed standard deviation of
observation error of $0.2$ on a log scale. Models in \textbf{B} and \textbf{D}
that had not met convergence thresholds were run for four chains, 256,000 iterations (128,000 warm up), and retaining every 10th sample, after which the MCMC chains for one population in \textbf{B} and four populations in \textbf{D} had still not met convergence thresholds. These populations are excluded from their respective panels, but all populations with Pr($\nu < 10$) $> 0.5$ in the base model had converged and are shown.}

@@

\caption{Estimates of $\nu$ from Gompertz models with alternative priors on $\nu$. \textbf{A} shows a weaker prior on $\nu$, \textbf{B} shows a stronger prior on $\nu$, as illustrated in Fig.~1. Shown are medians of the posterior (dots) and 50\% credible intervals (segments). The diagonal line indicates a one-to-one relationship. Different colours indicate various taxonomic classes. The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics. In general, the estimates are nearly identical in cases where the data are informative about low values of $\nu$. When the data are less informative about low values of $\nu$, the prior can slightly pull the estimates of $\nu$ towards higher or lower values.}

@@

\caption{Testing the ability to estimate $\nu$ (top panels) and the scale parameter of the process deviations (bottom panels) for a given number of samples (columns) drawn from a distribution with a given true $\nu$ value (rows). The red lines indicate the true population value. When a small number of samples are drawn there may not be samples sufficiently far into the tails to recapture the true $\nu$ value; however, heavy tails are still distinguished from normal tails in most cases, even with only 25 or 50 samples.}

@@

\caption{Simulation testing the Gompertz estimation model when the process deviation draws were chosen so that $\nu$ could be estimated close to the true value outside the full population model (``effective $\nu$'' within a CV of 0.2 of specified $\nu$). Upper panels show the distribution of median $\widehat{\nu}$ across 20 simulation runs. Lower panels show the distribution of Pr($\nu < 10$) across 20 simulation runs. We ran the simulations across three population (``true'') $\nu$ values (3, 5, and $1\cdot 10^9$, i.e.\ approximately normal) and four scenarios: (1) 100 time steps and no observation error, (2) 50 time steps and no observation error, (3) 50 time steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4) 50 time steps with observation error in which the quantity of observation error was assumed known. Within each scenario the dots represent stochastic draws from the true population distributions combined with model fits. Underlayed boxplots show the median, interquartile range, and $1.5$ times the interquartile range.}

@@

\caption{Potential covariates of heavy-tailed population dynamics (indicated by a high probability that $\nu < 10$). Shown are (\textbf{A} to \textbf{C}) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (\textbf{D}) number of time steps, (\textbf{E}) body length, and (\textbf{F}) lifespan. For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-$t$ process-noise distribution, $\lambda$ refers to the expected log abundance at the next time step at an abundance of one, $b$ refers to the density dependence parameter. Circles representing a few sharks, crustaceans, and gastropods are filled in white. Median and 90\% credible interval posterior predictions of a beta regression hierarchical model are shown in panels \textbf{A} and \textbf{D} where there was a high probability the slope coefficient was different from zero (Fig.~3B).}

@@

\caption{Skewness and heavy-tailed parameter posteriors. Top panel shows the shape of the skew-$t$ distribution at different combinations of skewness ($\gamma$) and heavy-tailedness ($\nu$). Lower panel shows the posterior medians and 50\% credible intervals for the two parameters. The horizontal red line indicates the median ln(skewness) parameter across all populations.}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1.1\textwidth]{../analysis/heavy-skew-projections-ggplot}
\caption{Projections of population abundance under skew-$t$ (red) and
normally distributed process error models (blue) for populations with
Pr($\nu<10$) $>0.5$. Numbers above panels indicate GPDD population IDs as
shown in table~S2. Horizontal axes indicates year before last
recorded abundance and a projection five years into the future. For the
projected region, solid lines represents the median projection and
dashed lines represent lower 99\% credible intervals. Vertical axes are
log-distributed. Assuming normal process error tends to underestimate risk of
low abundance if the process error is in fact heavy-tailed.}
